Linear Ordered Attribute Grammar scheduling using SAT-solving

### 2014-10-17 paper/paper.tex
Proofreading (improvements sections still missing)

\section{Introduction}
Attribute Grammars \cite{knuth68} are a formalism for describing the semantics of context-free languages, thereby making them suitable for the construction of compilers. Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler. The UHC is also the motivation for this paper and is used as a real world test case for testing the effectiveness of the given approach.

An attribute grammar essentially describes computations over trees, also known as folds or catamorphisms. Although the attribute grammar defines exactly \emph{what} should be computed, it does not define \emph{when} values should be computed. It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order. Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps. Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.

@@

\subsection{Overview}
In this paper we make the following main contributions:
\begin{itemize}
\item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT-problem
\item We show that chordal graphs can be used to encode cycle-freeness in SAT-problems involving graph encoding
\item We show how certain attribute grammar optimisations can be encode in the SAT-problem
\end{itemize}
Furthermore, we have implement the described techniques in the UUAGC\footnote{\url{http://www.cs.uu.nl/wiki/HUT/AttributeGrammarSystem}} \cite{combinator-languages} and show that the technique works well in practice for the UHC.

@@

In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the nonterminal \emph{Bin} with a complete order on the attributes. In this case there is only a single nonterminal so both the parent and the child nodes have the same attribute order, made explicit by extra edges from the nonterminal dependency graph. Because this dependency graph is still cycle free after adding the edges for the complete order and the same holds for the \emph{Leaf} production, the order a2 $\rightarrow$ a4 $\rightarrow$ a1 $\rightarrow$ a3 is a valid schedule for the nonterminal \emph{Tree}.

@@

@@

\begin{itemize}
\item \emph{Kastens}: the algorithm from \cite{kastens80} which only works for ordered attribute grammars, of which MainAG is not a member. However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously. However, as writing these annotation is hard an error-prone, this is not an optimal solution.
\item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linear ordered attribute grammars. However, the scheduling is not completely static so the generated evaluator also does part of the scheduling at runtime. 
\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linear ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm. This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper.
\end{itemize}

@@

\section{Optimisations} \label{sect:optimisations}
We have shown that expressing the scheduling problem as a SAT problem and using an existing SAT-solver can improve the running time of the scheduling, but that is not the only advantage. In the SAT problem one can easily add extra constraints to further influence the resulting schedule. In this section we show two of such optimisations that are useful from an attribute grammar perspective.

@@

\subsection{Eager attributes}
Another optimisation is the ability of defining \emph{eager attributes}. Eager attributes are attributes that should be computed as soon as possible, and must be annotated by the attribute grammar programmer as such. We would like our scheduling algorithm then to schedule them as early as possible in the total order.

@@

In the translation to the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs. This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions. We believe that this technique is applicable in other problems using SAT-solvers too.

@@

>>>>

\section{Introduction}
Attribute Grammars \cite{knuth68} are a formalism for describing the semantics of context-free languages, thereby making them suitable for the construction of compilers. Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler. The UHC is also the motivation for this paper and is used as a real world test case for evaluating the effectiveness of the given approach.

An attribute grammar essentially describes computations over trees, also known as folds or catamorphisms. Although the attribute grammar defines exactly \emph{what} should be computed, it does not define \emph{when} values should be computed. It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order. Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps. Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.

@@

\subsection{Overview}
In this paper we make the following main contributions:
\begin{itemize}
\item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT-problem
\item We show that chordal graphs can be used to encode cycle-freeness in SAT-problems involving graph encoding
\item We show how certain attribute grammar optimisations can be encoded in the SAT-problem
\end{itemize}
Furthermore, we have implement the described techniques in the UUAGC\footnote{\url{http://www.cs.uu.nl/wiki/HUT/AttributeGrammarSystem}} \cite{combinator-languages} and show that the technique works well in practice for the UHC.

@@

In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the production \emph{Bin} with a complete order on the attributes. In this case there is only a single nonterminal so both the parent and the child nodes have the same attribute order, made explicit by extra edges from the nonterminal dependency graph. Because this dependency graph is still cycle free after adding the edges for the complete order and the same holds for the \emph{Leaf} production, the order a2 $\rightarrow$ a4 $\rightarrow$ a1 $\rightarrow$ a3 is a valid schedule for the nonterminal \emph{Tree}.

@@

\subsection{Shared edges}
One important implementation detail is that of shared edges. As explained, the nonterminal dependency graphs and the production dependency graphs share the edges that define the order of the attributes. Because each edge is represented by a variable in the SAT problem we can simply encode this by assigning the same variable to the shared edges.

However, as we also make both graphs chordal, the edges added to make the graphs chordal can also be shared. This is exactly what our implementation does, such that the SAT-problem is kept as small as possible. The implementation is therefore slightly more complicated than explained in the previous sections.

@@

\begin{itemize}
\item \emph{Kastens}: the algorithm from \cite{kastens80} which only works for ordered attribute grammars, of which MainAG is not a member. However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously. However, as writing these annotation is hard an error-prone, this is not an optimal workflow.
\item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linear ordered attribute grammars. However, the scheduling is not completely static so the generated evaluator also does part of the scheduling at runtime. 
\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linear ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm. This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper and uses exponential time in worst-case.
\end{itemize}

@@

\section{Optimisations} \label{sect:optimisations}
We have shown that expressing the scheduling problem as a SAT problem and using an existing SAT-solver can improve the running time of the scheduling, but that is not the only advantage. In the SAT problem one can easily add extra constraints to further influence the resulting schedule. In this section we show two of such optimisations that are useful from an attribute grammar perspective. These optimisations have not been implemented in the release version of the UUAGC, but we have run preliminary experiments to verify that they work as expected.

@@

\subsection{Eager attributes}
Another optimisation is the ability to define \emph{eager attributes}. Eager attributes are attributes that should be computed as soon as possible, and must be annotated by the attribute grammar programmer as such. We would like our scheduling algorithm then to schedule them as early as possible in the total order.

@@

In the translation to the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs. This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions. We believe that this technique is applicable in other problems using SAT-solvers as well.

@@

All in all, we believe that this approach fully solves the basic scheduling problem in an elegant way. There are ample possibilities for improving the resulting schedules based on attribute grammar knowledge like the two discussed in \sectref{sect:optimisations}, so we have also made room for future improvements in the scheduling and compilation of attribute grammars.
